Human Factors Influencing Visual Statistical Inference

### 2013-06-02 sociology_paper.tex
factors affecting the peroformance updated

% Article top matter
\title{Impact of Demographics and Skills of the Observer on Visual Statistical Inference }
\author{{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\thanks{Mahbubul Majumder is a PhD student (e-mail: mahbub72@gmail.com) , Heike Hofmann is an Associate  Professor and Dianne Cook is a Professor in the Department of Statistics and Statistical Laboratory, Iowa State University, Ames, IA 50011-1210. This research is supported in part by the National Science Foundation Grant \# DMS 1007697.}}
\date{\vspace{-.5in}}
%\date{\today}  %\today is replaced with the current date
\maketitle

@@

\subsection{Visual Perception of Human Eye} Different people observe the lineup in different ways.  With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The result suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and then start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too.

Given this pattern of human eye tracks, it may effect the performance of an observer depending on where the actual plot is placed in the lineup. If the actual plot is on the top left corner and the observer starts from that point it may be easier to detect the actual plot earlier in the exercise and the observer could get plenty of time to make comparison. Thus the position of the actual plot in a lineup may have some impact.

@@

@@

\paragraph{Design for Location effect:} The experiment is set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots are obtained by randomizing the group structure of the data. The actual plot is then randomly placed in five different locations. For each location, five different sets of null plots are used to produce five different lineups. This produces 25 different lineups. One such lineup is shown in Figure \ref{fig:lineup_geno}.Similar procedure is applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

@@

\subsection{Learning Trend of the Observer} Each subject participating the experiment has seen multiple number of lineups for evaluations. Suppose $K$ be the number of lineups evaluated by a subject and $A_k$ be the $kth$ attempt, $k=1,2, ..., K$. Subjects may have a chance for self learning and do better on lineups shown later in the sequence. They may learn from their previous mistake or they may learn about the features of the plots being evaluated as they progress. Thus if there is any learning trend that should be apparent in proportion correct over $A_k$. But to obtain this we have to adjust for lineup difficulty and individual performances.

@@

\subsection{Location Effect of Data Plot in the Lineup}

@@

The another simulation experiment reveals that the effect of location of actual data plot in the lineup is not significant. This is important as the visual statistical inference procedure suggests that the data plot be placed at random anywhere in the lineup. This paper suggests that any random place in a lineup is as good as other places in the lineup.

More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference is higher with skilled observer.

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer. Also a lineup with fewer or larger than 10 plots may show different results. If the size of the lineup is much higher than 20, it is intuitive that there may be location effect of the data plot in the lineup.

>>>>

% Article top matter
\title{Human Factors Influencing Visual Statistical Inference }
\author{{Mahbubul Majumder, Heike Hofmann, Dianne Cook}
\thanks{Mahbubul Majumder is a PhD student (e-mail: mahbub72@gmail.com), Heike Hofmann is an Associate  Professor and Dianne Cook is a Professor in the Department of Statistics and Statistical Laboratory, Iowa State University, Ames, IA 50011-1210. This research is supported in part by the National Science Foundation Grant \# DMS 1007697.}}
\date{\vspace{-.5in}}
%\date{\today}  %\today is replaced with the current date
\maketitle

@@

%\subsection{Visual Perception of Human Eye} Different people observe the lineup in different ways.  With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The result suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and then start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too. 
%
%Given this pattern of human eye tracks, it may effect the performance of an observer depending on where the actual plot is placed in the lineup. If the actual plot is on the top left corner and the observer starts from that point it may be easier to detect the actual plot earlier in the exercise and the observer could get plenty of time to make comparison. Thus the position of the actual plot in a lineup may have some impact.

@@

\subsection{Learning Trend of the Observer} When an observer evaluates a lineup he or she may learn something from the experience. If the same observer is shown another lineup the learning from previous evaluation may help. The more evaluations an observer makes the more skillful the observer may become. This learning trend may or may not be significant overall but it could influence the performance of the observer in some way. This paper investigates the learning trend with multiple simulation experiments.

\subsection{Location of Actual Plot in the Lineup} The actual data plot is placed in a random spot in a lineup. While evaluating the lineup some people may start looking from some specific part of the lineup. With the help of eye-tracking equipment \citet{zhao:2012} track the observers' eyes to see how they go through the plots in a lineup to come to their answers. The results suggest that people have particular methods of reading lineups. Some people read lineups from left to write direction while some read from upward to downward. Some people start looking from the center of the lineup while others start from the top left corner. In the earlier phase of the exercise, the observer tend to scan the plots and then start comparing plots to make a final decision. Beside right-left or up-down directions observer show some diagonal movement too.

Given a specific pattern of eye movement of the observer in examining the lineup, the location of actual plot in the lineup may have some effect. Those who start exploring from the left top corner may get first glance of the actual plot if it is on that location. This should give the observer more time to compare it with the rest of the null plots. Those who start looking from the center may then scan towards right or left direction and thus don't have the chance to scan the null plots continuously as they could if they would start from a corner. Thus they should scan the center plot over and over again while scanning the left or right side plots. If the data plot is in the center location it may have multiple chance to be examined and hence be identified.

It may be possible that some part of the lineup is little explored or not scanned by the observer if he or she feels that the actual plot is found before even seriously scanning the whole lineup. In those scenarios the location where the observer first start scanning is important. Placing the actual data plot in that location may yield different results.

\subsection{Selection of Null Plots} A lineup becomes difficult to evaluate if one of the null plots appears to be very similar to the actual plot. When null hypothesis is true it is a common scenario. With alternative hypothesis being true,  it may happen if  we compare actual plot with many null plots. But we have only specific number of null plots in a lineup. So, a different set of null plots may yield some variation in the difficulty of the lineup with the same actual plot.

@@

\paragraph{Design for Location effect:} The experiment is set up based on the gene expression data with two groups. The groups are defined based on the presence or absence of iron sufficiency. The null hypothesis is that there is no difference between this two groups. Null plots are obtained by randomizing the group structure of the data. The actual plot is then randomly placed in five different locations. For each location, five different sets of null plots are used to produce five different lineups. This produces 25 different lineups. One such lineup is shown in Figure \ref{fig:lineup_geno}. Similar procedure is applied to obtain 25 lineups to test the interaction between Genotype and Empty Vector (EV).

@@

\subsection{Learning Trend} Each subject participating the experiment has seen multiple number of lineups for evaluations. Suppose $K$ be the number of lineups evaluated by a subject and $A_k$ be the $kth$ attempt, $k=1,2, ..., K$. Subjects may have a chance for self learning and do better on lineups shown later in the sequence. They may learn from their previous mistake or they may learn about the features of the plots being evaluated as they progress. Thus if there is any learning trend that should be apparent in proportion correct over $A_k$. But to obtain this we have to adjust for lineup difficulty and individual performances.

@@

\subsection{Location Effect}

@@

The simulation experiment reveals that the effect of location of actual data plot in the lineup is not significant. This is important as the visual statistical inference procedure suggests that the data plot be placed at random anywhere in the lineup. Even though there are variations on the performance depending on actual plot locations, it is not statistically significant to be of any concern. This paper suggests that any random place in a lineup is as good as other places in the lineup.

More experiments may be needed to make a concrete decision about the difference in power with skilled observer and a non-skilled observer in terms of statistical graphics. The numerous pilot studies done with the more trained participants suggest that power of visual statical inference may be higher with skilled observer. Thus the variations observed with individual performances could be attributed to individual abilities.

This investigation allowed each observer to evaluate 10 lineups assuming that it would not cause fatigue or disinterest of the observers towards the experiment based on the pilot studies. The future research may involve doing more experiments to check if fewer or more than 10 lineups have any significant impact on performance of the observer.

Also, a lineup with fewer or larger than 20 plots may yield different results. If the size of the lineup is much higher than 20, it is intuitive that there may be location effect of the data plot in the lineup. It is because, the observer may get tiered of scanning and make decision based on the partial scanning of the lineup. On the other hand fewer than 20 plots would give observer compare the actual plot with fewer null plots give more chance of picking actual plot as an error not as an actual plot. These are the some of the issues that needs to investigated.
